---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
### The data used for this assignment was obtained from [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) of the course website. To load the data iinto R, the following code was used
```{r Loading Data Into R}
actData <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, sep = ",")
```
#### Check the data
```{r Summary}
summary(actData)
```
```{r Class Check}
sapply(actData, class)
```
#### the date will be changed from 'factor' to 'date'
```{r Date class change}
actData$date <- as.Date(as.character(actData$date), "%Y-%m-%d")
```



## What is mean total number of steps taken per day?

#### To genarate the total steps taken each day, the code is as shown below was used
```{r Steps per day}
stpd <- aggregate(steps ~ date, actData, FUN = sum)
```

#### This is the code used to plot the histogram
```{r Histogram steps per day}
hist(stpd$steps, col = "dark green", main = " Histogram of the total number of steps taken each day ", 
     xlab = "Total No of Steps Each Day")
```

#### To check the summary amd obtain the mean and median
```{r mean & Median}
summary(stpd$steps)
```
#### The mean and median from the above summary is **10766** and **10765** respectively.



## What is the average daily activity pattern?

#### To genarate the five minutes interval and average steps taken each day, the code is as shown below was used
```{r 5-min interval average}
AvgInt <- aggregate(steps ~ interval, actData, FUN = mean)
```

#### Then a plot of the 5-minute interval on x-axis and the average number of steps taken, averaged across all days on y-axis is done
```{r Plot 5min int vs avg steps}
library(ggplot2)
ggplot(AvgInt, aes(interval, steps)) + geom_line() +xlab("5-Minutes Intervals") + ylab("Average Steps taken Across All Days")
```

#### To find the maximum steps on 5-minute intervals:
```{r Max no of step Interval}
maxi <- AvgInt$interval[which.max(AvgInt$steps)]
```
#### From the code above, maxi, the 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is **`r maxi`**


   
   
## Imputing missing values

#### To find the missing values the code 
```{r No of Missin values}
MisVal <- sum(is.na(actData$steps))
```
#### was used. Therefore, the total number of missing values in the dataset (i.e. the total number of rows with NAs) **`r MisVal`**.


#### The strategy i will use to fill the missing values is by replacing all the NA with the mean of the steps column. The mean is calculated below.
```{r mean of steps}
mean(actData$steps, na.rm = TRUE)
```

#### This is done done by creating a new dataset 'cactData' equal to the original but with the NA (missing data) filled in.

```{r create new dataset and replce NA}
cactData <- read.csv(unz("activity.zip", "activity.csv"), header = TRUE, sep = ",")
# to replace the NA with mean

cactData$steps[is.na(cactData$steps)] <- mean(cactData$steps, na.rm = TRUE)
summary(cactData)
```
```{r New Date class change}
cactData$date <- as.Date(as.character(cactData$date), "%Y-%m-%d")
```


#### Using the code below to genarate the total steps taken each day for the new dataset
```{r New Steps per day}
cstpd <- aggregate(steps ~ date, cactData, FUN = sum)
```


#### And finally to plot the new histogram with the code:
```{r New Histogram steps per day}
hist(cstpd$steps, col = "green", main = " Histogram of the total number of steps taken each day ", 
     xlab = "Total No of Steps Each Day")
```


#### To generate the mean and median of the two datasets and finding the difference
```{r New mean & Median}
mean1ds <- mean(stpd$steps)
median1ds <- median(stpd$steps)
mean2ds <- mean(cstpd$steps)
median2ds <- median(cstpd$steps)
meandiff <- mean2ds - mean1ds
mediandiff <- median2ds - median1ds
```


#### The new mean and median obtained is **`r mean2ds`** and **`r median2ds`** respectivelly.  
#### The difference between the two means is **`r meandiff`** while that of the median is **`r mediandiff`**.
   

#### By compairing the two histograms and adjusting the y-axis to be the same, impact of imputing missing data on the estimates of the total daily number of steps can be observed

```{r plot for compairing}
par(mfrow = c(1,2))
hist(cstpd$steps, col = "green", ylim =c(0,35))
hist(stpd$steps, col = " dark green", ylim =c(0,35))
```
   
   

## Are there differences in activity patterns between weekdays and weekends?

#### A new variable 'week' is created which indicates whether a given date is a weekday or weekend day
```{r create weekend or weekday column}
cactData$week <- ifelse(weekdays(cactData$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
head(cactData)
```


#### To make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). A new data for plotting us created.
```{r plotiing data}
AvgInt2 <- aggregate(steps ~ interval + week, cactData, FUN = mean)
```
#### The plotting code is shown below
```{r the plotting}
library(graphics)
library(lattice)
xyplot(steps ~ interval | week, AvgInt2, type = "l", 
       layout = c(1,2)
       )
```

